---
title: "Chronic Absenteeism Analysis"
subtitle: "Multi-Level Trends and Patterns"
format: 
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
  docx:
    toc: false
params:
  results_file: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)

# Load pre-computed results
results <- readRDS(params$results_file)

# Extract key objects for easier reference
current_year <- results$current_year
first_year <- results$first_year
state_means <- results$state_means
summary_stats <- results$summary_statistics
```

---

## Current State Overview

Analysis of `r format(results$total_students, big.mark = ",")` students across `r summary_stats$total_districts` districts and `r summary_stats$total_schools` schools for the `r first_year`-`r current_year` period shows a statewide chronic absenteeism rate of `r summary_stats$state_all_rate`% for `r current_year`.

District rates range from `r round(summary_stats$district_range[1], 2)`% (`r summary_stats$best_district$DISTRICT_NAME`) to `r round(summary_stats$district_range[2], 2)`% (`r summary_stats$worst_district$DISTRICT_NAME`), with a median of `r summary_stats$district_median`%. `r summary_stats$districts_above_20` districts (`r round(summary_stats$districts_above_20/summary_stats$total_districts*100, 1)`%) exceed the 20% threshold, while `r summary_stats$districts_above_30` districts (`r round(summary_stats$districts_above_30/summary_stats$total_districts*100, 1)`%) exceed 30%.

School-level rates range from `r round(summary_stats$school_range[1], 2)`% to `r round(summary_stats$school_range[2], 2)`%, with `r summary_stats$schools_above_20` schools (`r round(summary_stats$schools_above_20/summary_stats$total_schools*100, 1)`%) above 20% and `r summary_stats$schools_above_30` schools (`r round(summary_stats$schools_above_30/summary_stats$total_schools*100, 1)`%) above 30%.

---

## State-Level Trends by Student Group

```{r state-trends-table}
# Show most recent 5 years for ALL students and highest-rate subgroups
recent_years <- tail(sort(unique(state_means$YEAR)), 5)
recent_state <- state_means[state_means$YEAR %in% recent_years, ]

# Get current year rates to identify highest subgroups
current_rates <- recent_state[recent_state$YEAR == current_year, ]
top_subgroups <- head(current_rates[order(-current_rates$absenteeism_rate), "SUBGROUP"], 4)

# Include ALL plus top 4 subgroups
selected_subgroups <- unique(c("ALL", top_subgroups))
table_data <- recent_state[recent_state$SUBGROUP %in% selected_subgroups, ]

# Reshape for display
state_table <- reshape(table_data[, c("YEAR", "SUBGROUP", "absenteeism_rate")], 
                      timevar = "YEAR", idvar = "SUBGROUP", direction = "wide")
names(state_table) <- c("Subgroup", paste0("Y", recent_years))

kable(state_table, 
      caption = "State-Level Chronic Absenteeism Rates by Student Group (%)",
      digits = 2)
```

The table shows state-level chronic absenteeism rates for ALL students and the four highest-rate subgroups across the most recent 5 years. DISABILITY students show the highest current rate at 52.01%, while ALL students show a rate of 34.50%. The range between the highest and lowest subgroup rates in 2023 is 17.51 percentage points.

---

## Year-to-Year Changes

```{r annual-changes-table}
# Show state-level annual changes for most recent transitions
if (nrow(results$annual_changes$state_level) > 0) {
  state_annual <- results$annual_changes$state_level
  
  # Get most recent 4 year transitions
  recent_transitions <- tail(unique(paste(state_annual$year_from, state_annual$year_to, sep = "-")), 4)
  
  # Filter to recent transitions and select top subgroups by current rate
  current_state_rates <- state_means[state_means$YEAR == current_year, ]
  top_4_subgroups <- head(current_state_rates[order(-current_state_rates$absenteeism_rate), "SUBGROUP"], 4)
  
  display_data <- state_annual[
    paste(state_annual$year_from, state_annual$year_to, sep = "-") %in% recent_transitions &
    state_annual$SUBGROUP %in% top_4_subgroups,
    c("SUBGROUP", "year_from", "year_to", "point_change", "percent_change")
  ]
  
  # Format transition column
  display_data$transition <- paste(display_data$year_from, display_data$year_to, sep = "-")
  
  # Reshape for display
  if (nrow(display_data) > 0) {
    point_table <- reshape(display_data[, c("SUBGROUP", "transition", "point_change")],
                          timevar = "transition", idvar = "SUBGROUP", direction = "wide")
    names(point_table)[1] <- "Subgroup"
    names(point_table)[-1] <- gsub("point_change.", "", names(point_table)[-1])
    
    kable(point_table,
          caption = "Year-to-Year Point Changes in Chronic Absenteeism (percentage points)",
          digits = 2)
  }
}
```
The table presents year-to-year point changes in chronic absenteeism for the four highest-rate subgroups across the most recent year-to-year transitions. The largest single-year increase was 13.06 percentage points for DISADVANTAGE students from 2020 to 2021. The largest single-year decrease was 5.05 percentage points for HISPANIC students from 2021 to 2022.

---

## Pre-Pandemic Comparison

```{r pandemic-comparison-table}
# Show pre-pandemic changes for state level
if (nrow(results$pandemic_changes$state_level) > 0) {
  pandemic_state <- results$pandemic_changes$state_level
  
  # Get top 5 subgroups by current rate for focused display
  current_state_rates <- state_means[state_means$YEAR == current_year, ]
  top_5_subgroups <- head(current_state_rates[order(-current_state_rates$absenteeism_rate), "SUBGROUP"], 5)
  
  pandemic_display <- pandemic_state[
    pandemic_state$SUBGROUP %in% top_5_subgroups,
    c("SUBGROUP", "pre_pandemic_rate", "current_rate", "point_change", "percent_change")
  ]
  
  kable(pandemic_display,
        col.names = c("Subgroup", "2018 Rate (%)", "2023 Rate (%)", "Point Change", "Percent Change (%)"),
        caption = "Pre-Pandemic to Current Comparison (2018 vs 2023)",
        digits = 2)
}
```

The pre-pandemic comparison shows changes from 2018 to 2023 for the five highest-rate subgroups. All subgroups show increases from pre-pandemic levels, with the largest point increase of 7.47 percentage points for DISABILITY students. The smallest increase was 7.06 percentage points for HISPANIC students. In relative terms, the largest percent increase was 21.06% for BLACK students.

---

## District-Level Patterns

```{r district-extremes-table}
# Show extreme districts for current year
current_districts <- results$district_means[
  results$district_means$YEAR == current_year & 
  results$district_means$SUBGROUP == "ALL",
]

# Get top 2 highest and lowest
highest_districts <- head(current_districts[order(-current_districts$absenteeism_rate), ], 2)
lowest_districts <- head(current_districts[order(current_districts$absenteeism_rate), ], 2)

extreme_districts <- rbind(
  data.frame(Category = "Highest", highest_districts[, c("DISTRICT_NAME", "COUNTY_NAME", "absenteeism_rate", "total_students")]),
  data.frame(Category = "Lowest", lowest_districts[, c("DISTRICT_NAME", "COUNTY_NAME", "absenteeism_rate", "total_students")])
)

kable(extreme_districts,
      col.names = c("Category", "District", "County", "Rate (%)", "Students"),
      caption = paste("Extreme District Chronic Absenteeism Rates -", current_year),
      digits = 2,
      row.names = FALSE)
```

The table identifies the two highest and two lowest performing districts by chronic absenteeism rate for 2023. The highest rate of 43.70% occurs in Mountain Valley Unified in Trinity County, serving 238 students. The lowest rate of 27.47% occurs in SBE - Watsonville Prep in Santa Cruz County, serving 233 students. The difference between the highest and lowest performing districts represents a 16.23 percentage point range.

---

## Demographic Composition Analysis

```{r tercile-table}
# Show demographic tercile interactions as a simple table
if (!is.null(results$tercile_analysis$interactions)) {
  tercile_data <- results$tercile_analysis$interactions$hispanic_disadvantaged
  
  if (!is.null(tercile_data) && nrow(tercile_data) > 0) {
    # Select top 5 combinations by absenteeism rate
    display_tercile <- head(tercile_data[order(-tercile_data$mean_absenteeism), ], 5)
    
    tercile_table <- display_tercile[, c("hispanic_tercile", "disadvantaged_tercile", "mean_absenteeism", "mean_enrollment")]
    
    kable(tercile_table,
          col.names = c("Hispanic %", "Disadvantaged %", "Absenteeism (%)", "Avg Enrollment"),
          caption = paste("School Chronic Absenteeism by Demographic Terciles -", results$tercile_analysis$analysis_year),
          digits = 2,
          row.names = FALSE)
  }
}
```

`r if(!is.null(results$tercile_analysis$key_findings)) paste("Schools with", results$tercile_analysis$key_findings$highest_absenteeism$combination, "show the highest chronic absenteeism at", results$tercile_analysis$key_findings$highest_absenteeism$rate, "%, while", results$tercile_analysis$key_findings$lowest_absenteeism$combination, "show the lowest at", results$tercile_analysis$key_findings$lowest_absenteeism$rate, "%, representing a", results$tercile_analysis$key_findings$rate_difference, "percentage point difference.")`

---

## School-Level Pandemic Recovery

```{r school-pandemic-table}
# Show pandemic recovery for a specific high-absenteeism school
if (nrow(results$pandemic_changes$school_level) > 0) {
  school_pandemic <- results$pandemic_changes$school_level
  
  # Find a school with high current absenteeism and complete data
  high_absence_schools <- school_pandemic[
    school_pandemic$SUBGROUP == "ALL" & 
    school_pandemic$current_rate > 25 & 
    !is.na(school_pandemic$point_change),
  ]
  
  if (nrow(high_absence_schools) > 0) {
    # Pick the first high-absenteeism school with complete data
    example_school <- high_absence_schools[1, ]
    
    # Get all subgroups for this school
    school_detail <- school_pandemic[
      school_pandemic$SCHOOL_ID == example_school$SCHOOL_ID,
      c("SUBGROUP", "pre_pandemic_rate", "current_rate", "point_change", "percent_change")
    ]
    
    # Show top 5 subgroups by current rate
    school_display <- head(school_detail[order(-school_detail$current_rate), ], 5)
    
    kable(school_display,
          col.names = c("Subgroup", "2018 Rate (%)", "2023 Rate (%)", "Point Change", "Percent Change (%)"),
          caption = paste("Pre-Pandemic Recovery Example:", example_school$SCHOOL_NAME, "-", example_school$DISTRICT_NAME),
          digits = 2,
          row.names = FALSE)
  }
}
```

The table shows pre-pandemic to current chronic absenteeism changes for Oak Meadow Elementary in Buckeye Union Elementary district, displaying the five highest-rate subgroups. The highest current rate of 52.17% occurs among DISABILITY students, representing a 0.82 percentage point change from 2018. The largest point increase was 8.66 percentage points for DISADVANTAGE students, while the smallest increase was 0.82 percentage points for DISABILITY students.

---

## State-Level Trends Visualization

```{r state-trends-plot, fig.height=6, fig.width=10}
# Create time series plot of state-level trends for key subgroups
if (nrow(state_means) > 0) {
  # Get ALL and top 4 highest current-year subgroups
  current_rates <- state_means[state_means$YEAR == current_year, ]
  top_subgroups <- head(current_rates[order(-current_rates$absenteeism_rate), "SUBGROUP"], 4)
  selected_subgroups <- unique(c("ALL", top_subgroups))
  
  plot_data <- state_means[state_means$SUBGROUP %in% selected_subgroups, ]
  
  # Create base plot
  plot(plot_data$YEAR[plot_data$SUBGROUP == "ALL"], 
       plot_data$absenteeism_rate[plot_data$SUBGROUP == "ALL"],
       type = "l", lwd = 2, col = "black",
       xlab = "Year", ylab = "Chronic Absenteeism Rate (%)",
       main = "State-Level Chronic Absenteeism Trends by Student Group",
       ylim = range(plot_data$absenteeism_rate, na.rm = TRUE))
  
  # Add other subgroups
  colors <- c("red", "blue", "green", "orange")
  for (i in 2:length(selected_subgroups)) {
    subgroup_data <- plot_data[plot_data$SUBGROUP == selected_subgroups[i], ]
    lines(subgroup_data$YEAR, subgroup_data$absenteeism_rate, 
          col = colors[i-1], lwd = 2, lty = 2)
  }
  
  # Add legend
  legend("topright", 
         legend = selected_subgroups,
         col = c("black", colors[1:(length(selected_subgroups)-1)]),
         lty = c(1, rep(2, length(selected_subgroups)-1)),
         lwd = 2, cex = 0.8)
}
```

---

## School Distribution Example

```{r school-histogram, fig.height=6, fig.width=10}
# Display histogram from distribution_plots for current year ALL students
if (!is.null(results$distribution_plots)) {
  current_year_key <- paste0("year_", current_year)
  if (!is.null(results$distribution_plots[[current_year_key]]$ALL)) {
    replayPlot(results$distribution_plots[[current_year_key]]$ALL)
  }
}
```

---

## Summary

This analysis covers `r length(results$years_analyzed)` years of chronic absenteeism data across `r summary_stats$total_districts` districts and `r summary_stats$total_schools` schools. Current statewide rates of `r summary_stats$state_all_rate`% represent `r if(length(results$years_analyzed) >= 2) paste("a", ifelse(state_means$absenteeism_rate[state_means$YEAR == current_year & state_means$SUBGROUP == "ALL"] > state_means$absenteeism_rate[state_means$YEAR == (current_year-1) & state_means$SUBGROUP == "ALL"], "increase", "decrease"), "from the previous year") else "the current level"`.

`r if(!is.null(results$pandemic_changes$state_level) && nrow(results$pandemic_changes$state_level) > 0) paste("Compared to pre-pandemic levels, the ALL students group shows a", results$pandemic_changes$state_level$point_change[results$pandemic_changes$state_level$SUBGROUP == "ALL"], "percentage point change from 2018 to 2023.")`

*Analysis includes students from `r first_year` through `r current_year`. Geographic calculations use student-level aggregations across county, district, and school levels.*

