# State and Peer Comparisons

## State Average Comparisons

```{r state-comparison-calcs, echo=FALSE}
# Get state comparison data for current year
state_current <- params$district_data %>% 
  filter(YEAR == current_year, !is.na(STATE_RATE)) %>%
  mutate(
    GAP_FROM_STATE_PP = round(GAP_FROM_STATE * 100, 1),
    DISTRICT_RATE_PP = round(CHRONIC_RATE * 100, 1),
    STATE_RATE_PP = round(STATE_RATE * 100, 1)
  )

overall_gap <- state_current %>% filter(SUBGROUP == "ALL") %>% pull(GAP_FROM_STATE_PP)
district_overall_rate <- state_current %>% filter(SUBGROUP == "ALL") %>% pull(DISTRICT_RATE_PP)
state_overall_rate <- state_current %>% filter(SUBGROUP == "ALL") %>% pull(STATE_RATE_PP)

# Find subgroups with largest positive and negative gaps
subgroup_gaps <- state_current %>% 
  filter(SUBGROUP != "ALL") %>%
  arrange(desc(GAP_FROM_STATE_PP))

largest_positive_gap_subgroup <- subgroup_gaps$SUBGROUP[1]
largest_positive_gap <- subgroup_gaps$GAP_FROM_STATE_PP[1]
largest_negative_gap_subgroup <- subgroup_gaps$SUBGROUP[nrow(subgroup_gaps)]
largest_negative_gap <- subgroup_gaps$GAP_FROM_STATE_PP[nrow(subgroup_gaps)]

gap_range <- max(subgroup_gaps$GAP_FROM_STATE_PP) - min(subgroup_gaps$GAP_FROM_STATE_PP)

# Calculate additional metrics for expanded analysis
male_gap <- state_current %>% filter(SUBGROUP == "MALE") %>% pull(GAP_FROM_STATE_PP)
female_gap <- state_current %>% filter(SUBGROUP == "FEMALE") %>% pull(GAP_FROM_STATE_PP)
hispanic_gap <- state_current %>% filter(SUBGROUP == "HISPANIC") %>% pull(GAP_FROM_STATE_PP)
white_gap <- state_current %>% filter(SUBGROUP == "WHITE") %>% pull(GAP_FROM_STATE_PP)
ell_gap <- state_current %>% filter(SUBGROUP == "ELL") %>% pull(GAP_FROM_STATE_PP)
disadvantaged_gap <- state_current %>% filter(SUBGROUP == "DISADVANTAGE") %>% pull(GAP_FROM_STATE_PP)

male_district_rate <- state_current %>% filter(SUBGROUP == "MALE") %>% pull(DISTRICT_RATE_PP)
male_state_rate <- state_current %>% filter(SUBGROUP == "MALE") %>% pull(STATE_RATE_PP)
female_district_rate <- state_current %>% filter(SUBGROUP == "FEMALE") %>% pull(DISTRICT_RATE_PP)
female_state_rate <- state_current %>% filter(SUBGROUP == "FEMALE") %>% pull(STATE_RATE_PP)

# Count subgroups above and below state average
subgroups_above_state <- sum(subgroup_gaps$GAP_FROM_STATE_PP > 0)
subgroups_below_state <- sum(subgroup_gaps$GAP_FROM_STATE_PP < 0)
subgroups_equal_state <- sum(subgroup_gaps$GAP_FROM_STATE_PP == 0)
```

In `r current_year`, `r params$district_name` had an overall chronic absence rate of `r district_overall_rate`% compared to the state average of `r state_overall_rate`%.
This represents a gap of `r abs(overall_gap)` percentage points `r ifelse(overall_gap > 0, "above", "below")` the state average.

```{r state-gap-details, echo=FALSE, results='asis'}
if (overall_gap > 0) {
  cat("The district's chronic absence rate exceeded the state average.")
} else if (overall_gap < 0) {
  cat("The district's chronic absence rate was below the state average.")
} else {
  cat("The district's chronic absence rate matched the state average.")
}
```

Male students in the district had a chronic absence rate of `r male_district_rate`% compared to the state average of `r male_state_rate`% for male students.
This represents a gap of `r abs(male_gap)` percentage points `r ifelse(male_gap > 0, "above", "below")` the state average for this subgroup.
Female students had a rate of `r female_district_rate`% compared to the state average of `r female_state_rate`% for female students.
The gap for female students was `r abs(female_gap)` percentage points `r ifelse(female_gap > 0, "above", "below")` the state average.

Among demographic subgroups, `r largest_positive_gap_subgroup` students had the largest gap above the state average at `r largest_positive_gap` percentage points.
`r largest_negative_gap_subgroup` students had the gap furthest below the state average at `r largest_negative_gap` percentage points.
The range of gaps across subgroups spanned `r round(gap_range, 1)` percentage points.
Of the demographic subgroups analyzed, `r subgroups_above_state` had chronic absence rates above the state average for their subgroup.
`r subgroups_below_state` subgroups had rates below the state average.
`r if(subgroups_equal_state > 0) paste(subgroups_equal_state, "subgroups matched the state average exactly.") else "No subgroups matched the state average exactly."`

### District vs State Average Gap (Percentage Points)

```{r state-gaps-table, echo=FALSE}
# Gaps table - District rate minus state average
state_gaps <- params$district_data %>%
  filter(!is.na(STATE_RATE)) %>%
  select(YEAR, SUBGROUP, GAP_FROM_STATE) %>%
  mutate(GAP = round(GAP_FROM_STATE * 100, 1)) %>%
  select(YEAR, SUBGROUP, GAP) %>%
  arrange(match(SUBGROUP, c("ALL", "MALE", "FEMALE", "HISPANIC", "WHITE", "ASIAN", "BLACK", "ELL", "DISADVANTAGE", "DISABILITY")), YEAR) %>%
  tidyr::pivot_wider(names_from = YEAR, values_from = GAP)

kable(state_gaps, 
      caption = "Positive values indicate district chronic absence rate exceeds state average. Negative values indicate district rate is below state average.",
      align = c("l", rep("r", ncol(state_gaps)-1)))
```

## District Rankings

```{r ranking-calcs, echo=FALSE}
# Get ranking data for current year
ranking_current <- params$district_data %>%
  filter(YEAR == current_year, !is.na(DISTRICT_RANK)) %>%
  arrange(match(SUBGROUP, c("ALL", "MALE", "FEMALE", "HISPANIC", "WHITE", "ASIAN", "BLACK", "ELL", "DISADVANTAGE", "DISABILITY")))

overall_rank <- ranking_current %>% filter(SUBGROUP == "ALL") %>% pull(DISTRICT_RANK)
total_districts <- ranking_current %>% filter(SUBGROUP == "ALL") %>% pull(TOTAL_DISTRICTS)
overall_percentile <- ranking_current %>% filter(SUBGROUP == "ALL") %>% pull(DISTRICT_PERCENTILE)

best_ranking_subgroup <- ranking_current %>% filter(SUBGROUP != "ALL") %>% arrange(DISTRICT_RANK) %>% slice(1)
worst_ranking_subgroup <- ranking_current %>% filter(SUBGROUP != "ALL") %>% arrange(desc(DISTRICT_RANK)) %>% slice(1)

best_rank <- best_ranking_subgroup$DISTRICT_RANK
best_percentile <- best_ranking_subgroup$DISTRICT_PERCENTILE
worst_rank <- worst_ranking_subgroup$DISTRICT_RANK
worst_percentile <- worst_ranking_subgroup$DISTRICT_PERCENTILE

# Get additional ranking metrics
male_rank <- ranking_current %>% filter(SUBGROUP == "MALE") %>% pull(DISTRICT_RANK)
male_percentile <- ranking_current %>% filter(SUBGROUP == "MALE") %>% pull(DISTRICT_PERCENTILE)
female_rank <- ranking_current %>% filter(SUBGROUP == "FEMALE") %>% pull(DISTRICT_RANK)
female_percentile <- ranking_current %>% filter(SUBGROUP == "FEMALE") %>% pull(DISTRICT_PERCENTILE)

hispanic_rank <- ranking_current %>% filter(SUBGROUP == "HISPANIC") %>% pull(DISTRICT_RANK)
hispanic_percentile <- ranking_current %>% filter(SUBGROUP == "HISPANIC") %>% pull(DISTRICT_PERCENTILE)
white_rank <- ranking_current %>% filter(SUBGROUP == "WHITE") %>% pull(DISTRICT_RANK)
white_percentile <- ranking_current %>% filter(SUBGROUP == "WHITE") %>% pull(DISTRICT_PERCENTILE)

# Calculate districts with better/worse performance
districts_better <- overall_rank - 1
districts_worse <- total_districts - overall_rank

# Calculate ranking spread
ranking_spread <- worst_rank - best_rank
median_rank <- round(median(ranking_current %>% filter(SUBGROUP != "ALL") %>% pull(DISTRICT_RANK)), 0)
```

Among `r total_districts` districts statewide, `r params$district_name` ranked `r overall_rank` for overall chronic absence rate in `r current_year`.
This places the district at the `r overall_percentile`th percentile.
A ranking of `r overall_rank` means `r districts_better` districts had lower chronic absence rates.
`r districts_worse` districts had higher chronic absence rates than `r params$district_name`.

Male students in the district ranked `r male_rank` statewide, placing them at the `r male_percentile`th percentile among districts.
Female students ranked `r female_rank` statewide at the `r female_percentile`th percentile.
Hispanic students ranked `r hispanic_rank` with a percentile of `r hispanic_percentile`.
White students ranked `r white_rank` at the `r white_percentile`th percentile.

The district's best ranking among subgroups was for `r best_ranking_subgroup$SUBGROUP` students at rank `r best_rank` (percentile `r best_percentile`).
The district's highest ranking was for `r worst_ranking_subgroup$SUBGROUP` students at rank `r worst_rank` (percentile `r worst_percentile`).
Rankings ranged from `r best_rank` to `r worst_rank` across demographic subgroups.
This represents a spread of `r ranking_spread` positions between the best and worst performing subgroups.
The median ranking across all demographic subgroups was `r median_rank`.

### District Percentiles by Subgroup

```{r percentiles-table, echo=FALSE}
# Create percentiles table with years as columns
percentiles_wide <- params$district_data %>%
  filter(!is.na(DISTRICT_PERCENTILE)) %>%
  select(YEAR, SUBGROUP, DISTRICT_PERCENTILE) %>%
  arrange(match(SUBGROUP, c("ALL", "MALE", "FEMALE", "HISPANIC", "WHITE", "ASIAN", "BLACK", "ELL", "DISADVANTAGE", "DISABILITY")), YEAR) %>%
  tidyr::pivot_wider(names_from = YEAR, values_from = DISTRICT_PERCENTILE)

kable(percentiles_wide, 
      caption = "Lower percentiles indicate better chronic absence performance relative to all districts statewide.",
      align = c("l", rep("r", ncol(percentiles_wide)-1)))
```

## Peer District Comparisons

```{r peer-comparison-calcs, echo=FALSE}
# Get peer comparison data for current year
peer_current <- params$district_data %>%
  filter(YEAR == current_year, !is.na(PEER_RATE)) %>%
  mutate(
    GAP_FROM_PEERS_PP = round(GAP_FROM_PEERS * 100, 1),
    DISTRICT_RATE_PP = round(CHRONIC_RATE * 100, 1),
    PEER_RATE_PP = round(PEER_RATE * 100, 1)
  )

peer_size_band <- peer_current %>% filter(SUBGROUP == "ALL") %>% pull(PEER_SIZE_BAND)
peer_overall_gap <- peer_current %>% filter(SUBGROUP == "ALL") %>% pull(GAP_FROM_PEERS_PP)
district_peer_rate <- peer_current %>% filter(SUBGROUP == "ALL") %>% pull(DISTRICT_RATE_PP)
peer_average_rate <- peer_current %>% filter(SUBGROUP == "ALL") %>% pull(PEER_RATE_PP)

# Find subgroups with largest positive and negative peer gaps
peer_subgroup_gaps <- peer_current %>%
  filter(SUBGROUP != "ALL") %>%
  arrange(desc(GAP_FROM_PEERS_PP))

largest_positive_peer_gap_subgroup <- peer_subgroup_gaps$SUBGROUP[1]
largest_positive_peer_gap <- peer_subgroup_gaps$GAP_FROM_PEERS_PP[1]
largest_negative_peer_gap_subgroup <- peer_subgroup_gaps$SUBGROUP[nrow(peer_subgroup_gaps)]
largest_negative_peer_gap <- peer_subgroup_gaps$GAP_FROM_PEERS_PP[nrow(peer_subgroup_gaps)]

peer_gap_range <- max(peer_subgroup_gaps$GAP_FROM_PEERS_PP) - min(peer_subgroup_gaps$GAP_FROM_PEERS_PP)

# Get additional peer comparison metrics
male_peer_gap <- peer_current %>% filter(SUBGROUP == "MALE") %>% pull(GAP_FROM_PEERS_PP)
female_peer_gap <- peer_current %>% filter(SUBGROUP == "FEMALE") %>% pull(GAP_FROM_PEERS_PP)
hispanic_peer_gap <- peer_current %>% filter(SUBGROUP == "HISPANIC") %>% pull(GAP_FROM_PEERS_PP)
white_peer_gap <- peer_current %>% filter(SUBGROUP == "WHITE") %>% pull(GAP_FROM_PEERS_PP)
ell_peer_gap <- peer_current %>% filter(SUBGROUP == "ELL") %>% pull(GAP_FROM_PEERS_PP)
disadvantaged_peer_gap <- peer_current %>% filter(SUBGROUP == "DISADVANTAGE") %>% pull(GAP_FROM_PEERS_PP)

male_peer_district_rate <- peer_current %>% filter(SUBGROUP == "MALE") %>% pull(DISTRICT_RATE_PP)
male_peer_rate <- peer_current %>% filter(SUBGROUP == "MALE") %>% pull(PEER_RATE_PP)
female_peer_district_rate <- peer_current %>% filter(SUBGROUP == "FEMALE") %>% pull(DISTRICT_RATE_PP)
female_peer_rate <- peer_current %>% filter(SUBGROUP == "FEMALE") %>% pull(PEER_RATE_PP)

# Count subgroups above and below peer average
peer_subgroups_above <- sum(peer_subgroup_gaps$GAP_FROM_PEERS_PP > 0)
peer_subgroups_below <- sum(peer_subgroup_gaps$GAP_FROM_PEERS_PP < 0)
peer_subgroups_equal <- sum(peer_subgroup_gaps$GAP_FROM_PEERS_PP == 0)
```

`r params$district_name` is classified as a `r peer_size_band` district based on enrollment size.
Compared to other `r peer_size_band` districts, `r params$district_name` had a chronic absence rate of `r district_peer_rate`% versus the peer group average of `r peer_average_rate`% in `r current_year`.
This represents a gap of `r abs(peer_overall_gap)` percentage points `r ifelse(peer_overall_gap > 0, "above", "below")` the peer district average.

```{r peer-gap-details, echo=FALSE, results='asis'}
if (peer_overall_gap > 0) {
  cat("The district's chronic absence rate exceeded the peer district average.")
} else if (peer_overall_gap < 0) {
  cat("The district's chronic absence rate was below the peer district average.")
} else {
  cat("The district's chronic absence rate matched the peer district average.")
}
```

Male students in the district had a chronic absence rate of `r male_peer_district_rate`% compared to `r male_peer_rate`% among peer districts.
This gap of `r abs(male_peer_gap)` percentage points places male students `r ifelse(male_peer_gap > 0, "above", "below")` the peer average.
Female students had a rate of `r female_peer_district_rate`% compared to the peer average of `r female_peer_rate`%.
The gap for female students was `r abs(female_peer_gap)` percentage points `r ifelse(female_peer_gap > 0, "above", "below")` the peer average.

Among demographic subgroups, `r largest_positive_peer_gap_subgroup` students had the largest gap above the peer average at `r largest_positive_peer_gap` percentage points.
`r largest_negative_peer_gap_subgroup` students had the gap furthest below the peer average at `r largest_negative_peer_gap` percentage points.
The range of gaps across subgroups compared to peer districts spanned `r round(peer_gap_range, 1)` percentage points.
Of the demographic subgroups analyzed, `r peer_subgroups_above` had chronic absence rates above the peer district average.
`r peer_subgroups_below` subgroups had rates below the peer average.
`r if(peer_subgroups_equal > 0) paste(peer_subgroups_equal, "subgroups matched the peer average exactly.") else "No subgroups matched the peer average exactly."`

### District vs Peer Average Gap (Percentage Points)

```{r peer-gaps-table, echo=FALSE}
# Peer gaps table - District rate minus peer average
peer_gaps <- params$district_data %>%
  filter(!is.na(PEER_RATE)) %>%
  select(YEAR, SUBGROUP, GAP_FROM_PEERS) %>%
  mutate(GAP = round(GAP_FROM_PEERS * 100, 1)) %>%
  select(YEAR, SUBGROUP, GAP) %>%
  arrange(match(SUBGROUP, c("ALL", "MALE", "FEMALE", "HISPANIC", "WHITE", "ASIAN", "BLACK", "ELL", "DISADVANTAGE", "DISABILITY")), YEAR) %>%
  tidyr::pivot_wider(names_from = YEAR, values_from = GAP)

kable(peer_gaps, 
      caption = "Positive values indicate district chronic absence rate exceeds peer district average. Negative values indicate district rate is below peer average.",
      align = c("l", rep("r", ncol(peer_gaps)-1)))
```
