---
title: "Student Engagement Analysis: Assessment Participation and Chronic Absenteeism"
subtitle: "Multi-Year Trends for Data Quality and Accountability"
format: 
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
  docx:
    toc: false
params:
  results_file: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)
library(ggplot2)

# Load pre-computed results
results <- readRDS(params$results_file)

# Extract key objects for easier reference
current_year <- results$current_year
first_year <- results$first_year
ela_summary <- results$ela_summary
math_summary <- results$math_summary
yearly_trends_ela <- results$yearly_trends_ela
yearly_trends_math <- results$yearly_trends_math
yearly_absent_ela <- results$yearly_absent_ela
yearly_absent_math <- results$yearly_absent_math
participation_drivers <- results$participation_drivers
absenteeism_drivers <- results$absenteeism_drivers
largest_participation_gap <- results$largest_participation_gap
largest_absenteeism_gap <- results$largest_absenteeism_gap
worst_participation <- results$worst_participation
best_participation <- results$best_participation
worst_absenteeism <- results$worst_absenteeism
best_absenteeism <- results$best_absenteeism
```

---

## What is the current state of assessment participation and chronic absenteeism across California districts?

**This analysis examines student engagement patterns across `r ela_summary$total_districts` California districts serving `r format(ela_summary$total_enrollment, big.mark = ",")` students during the `r first_year`-`r current_year` period.** **Current statewide performance shows median district participation rates of `r ela_summary$median_participation`% (ELA) and `r math_summary$median_participation`% (Math), while chronic absenteeism has reached median levels of `r ela_summary$median_absenteeism`%.**

**Assessment participation directly impacts the validity of accountability determinations and federal compliance status.** **The statewide landscape shows considerable variation, with district participation rates ranging from `r worst_participation$participation_rate`% (`r worst_participation$DISTRICT_NAME`) to `r best_participation$participation_rate`% (`r best_participation$DISTRICT_NAME`), demonstrating a `r round(best_participation$participation_rate - worst_participation$participation_rate, 2)` percentage point range across districts.**

**Notably, `r ela_summary$pct_below_95`% of districts fall below the federal 95% participation threshold, and `r ela_summary$pct_above_20`% exceed the 20% chronic absenteeism benchmark.** **Current data shows district absenteeism rates ranging from `r best_absenteeism$chronic_absent_rate`% (`r best_absenteeism$DISTRICT_NAME`) to `r worst_absenteeism$chronic_absent_rate`% (`r worst_absenteeism$DISTRICT_NAME`), representing a `r round(worst_absenteeism$chronic_absent_rate - best_absenteeism$chronic_absent_rate, 2)` percentage point range across the state.**

---

## How have participation and absenteeism patterns changed over time?

### Multi-Year Participation Trends

```{r participation-trends-table}
# Create combined trends table
years <- sort(unique(c(yearly_trends_ela$YEAR, yearly_trends_math$YEAR)))

participation_trends_table <- data.frame(
  YEAR = years,
  median_participation_ELA = yearly_trends_ela$median_rate[match(years, yearly_trends_ela$YEAR)],
  median_participation_MATH = yearly_trends_math$median_rate[match(years, yearly_trends_math$YEAR)],
  pct_below_95_ELA = yearly_trends_ela$below_95_pct[match(years, yearly_trends_ela$YEAR)],
  pct_below_95_MATH = yearly_trends_math$below_95_pct[match(years, yearly_trends_math$YEAR)]
)

kable(participation_trends_table,
      col.names = c("Year", "ELA Median %", "Math Median %", "ELA % Below 95%", "Math % Below 95%"),
      caption = "District Assessment Participation Trends by Subject and Year")
```

**The trend data reveals that participation rates experienced their most significant disruption in 2020, with median participation dropping to unprecedented levels before beginning recovery in subsequent years.** **The federal compliance picture shows `r yearly_trends_ela$below_95_pct[yearly_trends_ela$YEAR == current_year]`% of districts currently failing to meet the 95% threshold in ELA, compared to `r yearly_trends_ela$below_95_pct[yearly_trends_ela$YEAR == first_year]`% in `r first_year`.**

### Multi-Year Absenteeism Trends

```{r absenteeism-trends-table}
absenteeism_trends_table <- data.frame(
  YEAR = years,
  median_absenteeism_ELA = yearly_absent_ela$median_rate[match(years, yearly_absent_ela$YEAR)],
  median_absenteeism_MATH = yearly_absent_math$median_rate[match(years, yearly_absent_math$YEAR)],
  pct_above_20_ELA = yearly_absent_ela$above_20_pct[match(years, yearly_absent_ela$YEAR)],
  pct_above_20_MATH = yearly_absent_math$above_20_pct[match(years, yearly_absent_math$YEAR)]
)

kable(absenteeism_trends_table,
      col.names = c("Year", "ELA Median %", "Math Median %", "ELA % Above 20%", "Math % Above 20%"),
      caption = "District Chronic Absenteeism Trends by Subject and Year")
```

**Chronic absenteeism patterns show dramatic increases during the 2020-2021 period, with median district rates reaching peak levels before beginning to moderate in recent years.** **The proportion of districts exceeding the 20% threshold increased substantially during this period, with `r yearly_absent_ela$above_20_pct[yearly_absent_ela$YEAR == current_year]`% currently above this benchmark compared to `r yearly_absent_ela$above_20_pct[yearly_absent_ela$YEAR == first_year]`% in `r first_year`.**

---

## Are engagement challenges concentrated in specific schools within districts or distributed system-wide?

### School-Level Participation Drivers

```{r participation-drivers-plot, fig.height=6, fig.width=12}
if (nrow(participation_drivers) > 0) {
  # Filter and get top districts
  ela_viz <- participation_drivers[
    participation_drivers$total_schools >= 5 & 
    participation_drivers$SUBJECT == "ELA", 
  ]
  math_viz <- participation_drivers[
    participation_drivers$total_schools >= 5 & 
    participation_drivers$SUBJECT == "MATH", 
  ]
  
  if (nrow(ela_viz) > 0) {
    ela_top <- ela_viz[order(-ela_viz$participation_range), ][1:min(10, nrow(ela_viz)), ]
  } else {
    ela_top <- data.frame()
  }
  
  if (nrow(math_viz) > 0) {
    math_top <- math_viz[order(-math_viz$participation_range), ][1:min(10, nrow(math_viz)), ]
  } else {
    math_top <- data.frame()
  }
  
  participation_viz_data <- rbind(ela_top, math_top)
  
  if (nrow(participation_viz_data) > 0) {
    participation_viz_data$district_label <- paste(
      participation_viz_data$DISTRICT_NAME, 
      participation_viz_data$COUNTY_NAME, 
      sep = "\n"
    )
    
    ggplot(participation_viz_data, aes(x = reorder(district_label, participation_range))) +
      geom_col(aes(y = participation_range, fill = SUBJECT), alpha = 0.7) +
      geom_text(aes(y = participation_range / 2, 
                    label = paste0(schools_below_90_participation, "/", total_schools)), 
                size = 2.5, hjust = 0.5, vjust = 0.5, color = "white", fontface = "bold") +
      scale_fill_manual(values = c("ELA" = "steelblue", "MATH" = "darkgreen")) +
      coord_flip() +
      facet_wrap(~ SUBJECT, scales = "free", ncol = 2) +
      labs(title = "Districts with Largest School-to-School Participation Variation by Subject",
           subtitle = "Top 10 districts by participation range within each subject",
           x = "District", 
           y = "Participation Rate Range (percentage points)",
           caption = "Labels show number of schools below 90% participation") +
      theme_minimal() +
      theme(axis.text.y = element_text(size = 7),
            legend.position = "none")
  }
}
```

**This analysis identifies districts where participation challenges are concentrated in specific schools rather than distributed system-wide.** **Among districts with at least 5 schools, `r results$ela_driven_participation` show ELA participation problems driven by one-third or fewer of their schools, while `r results$math_driven_participation` show similar patterns in MATH, suggesting that targeted interventions at the school level could significantly improve district-wide compliance rates.**

### School-Level Absenteeism Drivers

```{r absenteeism-drivers-plot, fig.height=6, fig.width=10}
if (nrow(absenteeism_drivers) > 0) {
  absenteeism_viz_data <- absenteeism_drivers[absenteeism_drivers$total_schools >= 5, ]
  
  if (nrow(absenteeism_viz_data) > 0) {
    absenteeism_viz_data <- absenteeism_viz_data[
      order(-absenteeism_viz_data$absenteeism_range), 
    ][1:min(15, nrow(absenteeism_viz_data)), ]
    
    ggplot(absenteeism_viz_data, aes(x = reorder(paste(DISTRICT_NAME, COUNTY_NAME, sep = "\n"), absenteeism_range))) +
      geom_col(aes(y = absenteeism_range), fill = "coral", alpha = 0.7) +
      geom_text(aes(y = absenteeism_range / 2, 
                    label = paste0(schools_above_30_absenteeism, "/", total_schools)), 
                size = 3, hjust = 0.5, vjust = 0.5, color = "white", fontface = "bold") +
      coord_flip() +
      labs(title = "Districts with Largest School-to-School Absenteeism Variation",
           subtitle = "Range between highest and lowest performing schools within district",
           x = "District", 
           y = "Chronic Absenteeism Range (percentage points)",
           caption = "Labels show number of schools above 30% chronic absenteeism") +
      theme_minimal() +
      theme(axis.text.y = element_text(size = 8))
  }
}
```

**Similar to participation patterns, chronic absenteeism challenges are often concentrated within specific schools rather than being district-wide phenomena.** **Analysis reveals that `r results$districts_driven_absenteeism` districts have absenteeism concerns driven by one-third or fewer of their schools, indicating that school-specific factors—such as leadership, climate, or targeted populations—may be primary drivers of attendance challenges.**

**A critical finding is that participation and absenteeism challenges are often concentrated in a small number of schools within districts—`r results$ela_driven_participation` districts have ELA participation issues driven by one-third or fewer of their schools, `r results$math_driven_participation` districts show similar patterns for MATH participation, while `r results$districts_driven_absenteeism` districts show concentration patterns for chronic absenteeism.** This concentration suggests that targeted school-level interventions may be more effective than district-wide approaches in many cases.

---

## Which student groups show the largest engagement gaps compared to their district averages?

**Disaggregated data analysis reveals substantial variation in engagement patterns across student groups, with some populations experiencing significantly lower participation rates and higher absenteeism than their district averages.** **The largest recorded participation gap shows `r largest_participation_gap$DISAGGREGATED_GROUP` students in `r largest_participation_gap$DISTRICT_NAME` participating at `r round(largest_participation_gap$participation_rate, 2)`%, compared to a district average of `r round(largest_participation_gap$district_participation, 2)`%—a difference of `r round(largest_participation_gap$participation_gap, 2)` percentage points.**

**For chronic absenteeism, the most substantial gap appears among `r largest_absenteeism_gap$DISAGGREGATED_GROUP` students in `r largest_absenteeism_gap$DISTRICT_NAME`, with an absenteeism rate of `r round(largest_absenteeism_gap$chronic_absent_rate, 2)`% compared to the district average of `r round(largest_absenteeism_gap$district_absenteeism, 2)`%—representing a `r round(largest_absenteeism_gap$absenteeism_gap, 2)` percentage point difference.**

**Across all districts and student groups, `r results$significant_participation_gaps` instances show participation gaps of 15 percentage points or more, while `r results$significant_absenteeism_gaps` instances demonstrate similar magnitude gaps in chronic absenteeism.** These patterns suggest that engagement challenges disproportionately affect specific student populations, with implications for both educational equity and data quality in accountability systems.







---

## How complete is the recovery from pandemic disruption?

**Analysis of districts with complete data spanning the 2018-19 pre-pandemic baseline through the current 2023-24 school year reveals `r nrow(results$pandemic_recovery_ela$participation_details)` districts have sufficient longitudinal data for recovery assessment.** **Among these districts, assessment participation shows an average change of `r round(mean(results$pandemic_recovery_ela$participation_details$participation_change_since_pandemic, na.rm = TRUE), 1)` percentage points since the 2020-21 pandemic year, while chronic absenteeism has changed by an average of `r round(mean(results$pandemic_recovery_ela$absenteeism_details$absenteeism_change_since_pandemic, na.rm = TRUE), 1)` percentage points over the same period.**

**The distribution of participation recovery shows substantial variation across districts: `r sum(results$pandemic_recovery_ela$participation_details$participation_change_since_pandemic > 0, na.rm = TRUE)` districts have improved their participation rates since the pandemic disruption, while `r sum(results$pandemic_recovery_ela$participation_details$participation_change_since_pandemic < 0, na.rm = TRUE)` districts continue to show participation rates below their 2020-21 levels.** **The most improved district has gained `r round(max(results$pandemic_recovery_ela$participation_details$participation_change_since_pandemic, na.rm = TRUE), 1)` percentage points in participation since the pandemic year, while the most declined district has lost `r round(abs(min(results$pandemic_recovery_ela$participation_details$participation_change_since_pandemic, na.rm = TRUE)), 1)` percentage points.**

```{r participation-recovery-table}
# Show top 3 most improved and bottom 2 least improved
part_improved <- head(results$pandemic_recovery_ela$participation_details, 3)
part_declined <- tail(results$pandemic_recovery_ela$participation_details, 2)
part_table <- rbind(part_improved, part_declined)[, c("DISTRICT_NAME", "participation_pre_pandemic", "participation_pandemic", "participation_current", "participation_change_since_pandemic")]

kable(part_table,
      col.names = c("District", "Pre-Pandemic %", "Pandemic %", "Current %", "Change Since Pandemic"),
      caption = "Assessment Participation: Most and Least Recovered Districts")
```

**For chronic absenteeism, recovery patterns reveal that `r sum(results$pandemic_recovery_ela$absenteeism_details$absenteeism_change_since_pandemic < 0, na.rm = TRUE)` districts have successfully reduced absenteeism below their pandemic-era levels, while `r sum(results$pandemic_recovery_ela$absenteeism_details$absenteeism_change_since_pandemic > 0, na.rm = TRUE)` districts show absenteeism rates that remain above or have worsened beyond their 2020-21 crisis levels.** **The most successful district has reduced chronic absenteeism by `r round(abs(min(results$pandemic_recovery_ela$absenteeism_details$absenteeism_change_since_pandemic, na.rm = TRUE)), 1)` percentage points since the pandemic year, while the least successful has seen absenteeism increase by `r round(max(results$pandemic_recovery_ela$absenteeism_details$absenteeism_change_since_pandemic, na.rm = TRUE), 1)` percentage points.**

```{r absenteeism-recovery-table}
# Show top 3 most improved and bottom 2 least improved  
absent_improved <- head(results$pandemic_recovery_ela$absenteeism_details, 3)
absent_declined <- tail(results$pandemic_recovery_ela$absenteeism_details, 2)
absent_table <- rbind(absent_improved, absent_declined)[, c("DISTRICT_NAME", "absenteeism_pre_pandemic", "absenteeism_pandemic", "absenteeism_current", "absenteeism_change_since_pandemic")]

kable(absent_table,
      col.names = c("District", "Pre-Pandemic %", "Pandemic %", "Current %", "Change Since Pandemic"),
      caption = "Chronic Absenteeism: Most and Least Recovered Districts")
```

**Comparing current performance to pre-pandemic baselines reveals that `r sum(results$pandemic_recovery_ela$participation_details$participation_change_since_pre > -2, na.rm = TRUE)` districts operate within two percentage points of their 2018-19 participation levels, while `r sum(results$pandemic_recovery_ela$absenteeism_details$absenteeism_change_since_pre < 2, na.rm = TRUE)` districts have restored absenteeism to within two percentage points of pre-pandemic norms.**

---









## What are the implications for data quality?

**The concentration of engagement challenges within specific schools and student groups has important implications for data interpretation and intervention planning.** **When participation or absenteeism problems are driven by a small number of schools within a district, district-wide averages may mask the severity of issues at particular sites while simultaneously understating the performance of schools that are functioning effectively.**

**The school-level concentration patterns identified in this analysis suggest that `r results$ela_driven_participation + results$math_driven_participation + results$districts_driven_absenteeism` district-metric combinations could benefit from targeted, school-specific interventions rather than broad district-wide approaches.** This finding is particularly relevant for states developing technical assistance and support strategies, as resources may be more effectively deployed when focused on the specific schools driving district-level challenges.

**Student group disparities further compound these data quality concerns, as substantial gaps in participation and attendance may result in certain populations being under-represented in accountability measures.** **The `r results$significant_participation_gaps + results$significant_absenteeism_gaps` instances of large engagement gaps identified in this analysis represent student groups whose academic performance may not be accurately reflected in state accountability systems due to low assessment participation or chronic absence patterns.**

*This analysis encompasses data from `r first_year` through `r current_year` across `r ela_summary$total_districts` districts meeting minimum enrollment criteria. District-level calculations exclude County Offices of Education and require minimum enrollment of 30 students.*
