# District Overview

## Basic Enrollment and Chronic Absence Metrics

**Current Year: `r current_year`**

In `r current_year`, `r params$district_name` enrolled `r district_current %>% filter(SUBGROUP == "ALL") %>% pull(ENROLLMENT)` students across `r district_current %>% filter(SUBGROUP == "ALL") %>% pull(N_SCHOOLS)` schools.
Of these students, `r district_current %>% filter(SUBGROUP == "ALL") %>% pull(CHRONIC_COUNT)` were chronically absent, representing a district-wide chronic absence rate of `r round((district_current %>% filter(SUBGROUP == "ALL") %>% pull(CHRONIC_RATE)) * 100, 1)`%.

```{r enrollment-breakdown, echo=FALSE}
male_enrollment <- district_current %>% filter(SUBGROUP == "MALE") %>% pull(ENROLLMENT)
female_enrollment <- district_current %>% filter(SUBGROUP == "FEMALE") %>% pull(ENROLLMENT)
hispanic_enrollment <- district_current %>% filter(SUBGROUP == "HISPANIC") %>% pull(ENROLLMENT)
white_enrollment <- district_current %>% filter(SUBGROUP == "WHITE") %>% pull(ENROLLMENT)
ell_enrollment <- district_current %>% filter(SUBGROUP == "ELL") %>% pull(ENROLLMENT)
disadvantaged_enrollment <- district_current %>% filter(SUBGROUP == "DISADVANTAGE") %>% pull(ENROLLMENT)

male_chronic <- district_current %>% filter(SUBGROUP == "MALE") %>% pull(CHRONIC_COUNT)
female_chronic <- district_current %>% filter(SUBGROUP == "FEMALE") %>% pull(CHRONIC_COUNT)
hispanic_chronic <- district_current %>% filter(SUBGROUP == "HISPANIC") %>% pull(CHRONIC_COUNT)
white_chronic <- district_current %>% filter(SUBGROUP == "WHITE") %>% pull(CHRONIC_COUNT)
ell_chronic <- district_current %>% filter(SUBGROUP == "ELL") %>% pull(CHRONIC_COUNT)
disadvantaged_chronic <- district_current %>% filter(SUBGROUP == "DISADVANTAGE") %>% pull(CHRONIC_COUNT)

male_rate <- round((district_current %>% filter(SUBGROUP == "MALE") %>% pull(CHRONIC_RATE)) * 100, 1)
female_rate <- round((district_current %>% filter(SUBGROUP == "FEMALE") %>% pull(CHRONIC_RATE)) * 100, 1)
hispanic_rate <- round((district_current %>% filter(SUBGROUP == "HISPANIC") %>% pull(CHRONIC_RATE)) * 100, 1)
white_rate <- round((district_current %>% filter(SUBGROUP == "WHITE") %>% pull(CHRONIC_RATE)) * 100, 1)
ell_rate <- round((district_current %>% filter(SUBGROUP == "ELL") %>% pull(CHRONIC_RATE)) * 100, 1)
disadvantaged_rate <- round((district_current %>% filter(SUBGROUP == "DISADVANTAGE") %>% pull(CHRONIC_RATE)) * 100, 1)

total_enrollment <- district_current %>% filter(SUBGROUP == "ALL") %>% pull(ENROLLMENT)
```

The district's student population includes `r male_enrollment` male students (`r round((male_enrollment/total_enrollment)*100, 1)`% of total enrollment) and `r female_enrollment` female students (`r round((female_enrollment/total_enrollment)*100, 1)`% of total enrollment).
Among male students, `r male_chronic` were chronically absent, yielding a chronic absence rate of `r male_rate`%.
Female students had `r female_chronic` chronically absent, resulting in a chronic absence rate of `r female_rate`%.

Hispanic students represent `r hispanic_enrollment` students (`r round((hispanic_enrollment/total_enrollment)*100, 1)`% of district enrollment) with `r hispanic_chronic` chronically absent students and a chronic absence rate of `r hispanic_rate`%.
White students number `r white_enrollment` (`r round((white_enrollment/total_enrollment)*100, 1)`% of enrollment) with `r white_chronic` chronically absent and a rate of `r white_rate`%.

English language learners comprise `r ell_enrollment` students (`r round((ell_enrollment/total_enrollment)*100, 1)`% of total enrollment).
Among this population, `r ell_chronic` students were chronically absent, producing a chronic absence rate of `r ell_rate`%.
Economically disadvantaged students total `r disadvantaged_enrollment` (`r round((disadvantaged_enrollment/total_enrollment)*100, 1)`% of enrollment) with `r disadvantaged_chronic` chronically absent students and a rate of `r disadvantaged_rate`%.

The average school in the district enrolled `r round(total_enrollment / (district_current %>% filter(SUBGROUP == "ALL") %>% pull(N_SCHOOLS)), 0)` students.
Chronic absence affected students across all demographic groups, with rates ranging from `r round((district_current %>% filter(SUBGROUP != "ALL") %>% pull(CHRONIC_RATE) %>% min()) * 100, 1)`% to `r round((district_current %>% filter(SUBGROUP != "ALL") %>% pull(CHRONIC_RATE) %>% max()) * 100, 1)`% across subgroups.
These patterns reflect both the scale and distribution of chronic absence challenges within `r params$district_name`.

### District Totals by Subgroup

```{r district-totals, echo=FALSE}
district_totals <- district_current %>%
  select(SUBGROUP, ENROLLMENT, CHRONIC_COUNT, CHRONIC_RATE) %>%
  mutate(CHRONIC_RATE = round(CHRONIC_RATE * 100, 1)) %>%
  arrange(match(SUBGROUP, c("ALL", "MALE", "FEMALE", "HISPANIC", "WHITE", "ASIAN", "BLACK", "ELL", "DISADVANTAGE", "DISABILITY")))

kable(district_totals, 
      col.names = c("Subgroup", "Enrollment", "Chronic Count", "Chronic Rate (%)"),
      align = c("l", "r", "r", "r"))
```

### School-Level Variance

```{r school-variance-calcs, echo=FALSE}
all_variance <- district_current %>% filter(SUBGROUP == "ALL")
min_rate <- round(all_variance$SCHOOL_RATE_MIN * 100, 1)
max_rate <- round(all_variance$SCHOOL_RATE_MAX * 100, 1)
sd_rate <- round(all_variance$SCHOOL_RATE_SD * 100, 2)
q25_rate <- round(all_variance$SCHOOL_RATE_Q25 * 100, 1)
q75_rate <- round(all_variance$SCHOOL_RATE_Q75 * 100, 1)
n_schools <- all_variance$N_SCHOOLS

# Calculate additional metrics
rate_range <- max_rate - min_rate
iqr_range <- q75_rate - q25_rate

# Get subgroup with highest and lowest variance
subgroup_variance <- district_current %>%
  filter(SUBGROUP != "ALL") %>%
  arrange(desc(SCHOOL_RATE_SD))

highest_var_subgroup <- subgroup_variance$SUBGROUP[1]
highest_var_sd <- round(subgroup_variance$SCHOOL_RATE_SD[1] * 100, 2)
lowest_var_subgroup <- subgroup_variance$SUBGROUP[nrow(subgroup_variance)]
lowest_var_sd <- round(subgroup_variance$SCHOOL_RATE_SD[nrow(subgroup_variance)] * 100, 2)

# Schools above/below district average
district_avg <- round((district_current %>% filter(SUBGROUP == "ALL") %>% pull(CHRONIC_RATE)) * 100, 1)
schools_above_avg <- round((q75_rate - district_avg) / (max_rate - min_rate) * n_schools, 0)
schools_below_avg <- n_schools - schools_above_avg
```

Chronic absence rates varied across the `r n_schools` schools in the district.
The school with the lowest chronic absence rate recorded `r min_rate`% while the highest rate was `r max_rate`%.
This represents a range of `r rate_range` percentage points between the lowest and highest performing schools.
The standard deviation of school rates was `r sd_rate` percentage points across all schools.

The first quartile of schools had chronic absence rates at or below `r q25_rate`%.
The third quartile threshold was `r q75_rate`%.
The interquartile range spanned `r iqr_range` percentage points, indicating the spread of rates among the middle 50% of schools.
Half of all schools fell between these quartile boundaries.

Among demographic subgroups, `r highest_var_subgroup` students showed the highest variance across schools with a standard deviation of `r highest_var_sd` percentage points.
`r lowest_var_subgroup` students demonstrated the lowest variance with a standard deviation of `r lowest_var_sd` percentage points.
This difference indicates that some subgroups experience more uniform chronic absence patterns across schools than others.

The district-wide average chronic absence rate was `r district_avg`%.
Schools in the upper quartile exceeded this average by `r round(q75_rate - district_avg, 1)` percentage points.
Schools in the lower quartile fell below the average by `r round(district_avg - q25_rate, 1)` percentage points.
The gap between high-performing and low-performing schools represents `r round((max_rate - min_rate) / district_avg * 100, 0)`% of the district average rate.

These patterns demonstrate the distribution of chronic absence across the district's school system.
Some schools serve student populations with chronic absence rates well below the district average.
Other schools face chronic absence rates that exceed the district norm by substantial margins.

```{r district-variance, echo=FALSE}
district_variance <- district_current %>%
  select(SUBGROUP, SCHOOL_RATE_SD, SCHOOL_RATE_MIN, SCHOOL_RATE_MAX, 
         SCHOOL_RATE_Q25, SCHOOL_RATE_Q75, N_SCHOOLS) %>%
  mutate(
    SCHOOL_RATE_SD = round(SCHOOL_RATE_SD * 100, 2),
    SCHOOL_RATE_MIN = round(SCHOOL_RATE_MIN * 100, 1),
    SCHOOL_RATE_MAX = round(SCHOOL_RATE_MAX * 100, 1),
    SCHOOL_RATE_Q25 = round(SCHOOL_RATE_Q25 * 100, 1),
    SCHOOL_RATE_Q75 = round(SCHOOL_RATE_Q75 * 100, 1)
  ) %>%
  arrange(match(SUBGROUP, c("ALL", "MALE", "FEMALE", "HISPANIC", "WHITE", "ASIAN", "BLACK", "ELL", "DISADVANTAGE", "DISABILITY")))

kable(district_variance,
      col.names = c("Subgroup", "SD (%)", "Min (%)", "Max (%)", "Q25 (%)", "Q75 (%)", "Schools"),
      align = c("l", "r", "r", "r", "r", "r", "r"))
```

### Year-over-Year Changes

```{r yoy-detailed-calcs, echo=FALSE, results='asis'}
# Get most recent year change data for all subgroups
recent_changes_all <- params$district_data %>%
  filter(!is.na(CHRONIC_RATE_CHANGE)) %>%
  arrange(desc(YEAR)) %>%
  group_by(SUBGROUP) %>%
  slice(1) %>%
  ungroup()

# Overall district changes
overall_changes <- recent_changes_all %>% filter(SUBGROUP == "ALL")

if (nrow(overall_changes) > 0) {
  enrollment_change <- overall_changes$ENROLLMENT_CHANGE
  chronic_count_change <- overall_changes$CHRONIC_COUNT_CHANGE
  rate_change <- round(overall_changes$CHRONIC_RATE_CHANGE * 100, 1)
  prev_year <- overall_changes$YEAR - 1
  current_year_yoy <- overall_changes$YEAR
  prev_enrollment <- overall_changes$ENROLLMENT - enrollment_change
  prev_chronic_count <- overall_changes$CHRONIC_COUNT - chronic_count_change
  prev_rate <- round((overall_changes$CHRONIC_RATE - overall_changes$CHRONIC_RATE_CHANGE) * 100, 1)
  current_enrollment <- overall_changes$ENROLLMENT
  current_chronic_count <- overall_changes$CHRONIC_COUNT
  current_rate <- round(overall_changes$CHRONIC_RATE * 100, 1)
  
  # Subgroup changes
  subgroup_changes <- recent_changes_all %>% 
    filter(SUBGROUP %in% c("MALE", "FEMALE", "HISPANIC", "WHITE", "ELL", "DISADVANTAGE")) %>%
    mutate(rate_change_pp = round(CHRONIC_RATE_CHANGE * 100, 1)) %>%
    arrange(desc(rate_change_pp))
  
  largest_increase_subgroup <- subgroup_changes$SUBGROUP[1]
  largest_increase_change <- subgroup_changes$rate_change_pp[1]
  largest_decrease_subgroup <- subgroup_changes$SUBGROUP[nrow(subgroup_changes)]
  largest_decrease_change <- subgroup_changes$rate_change_pp[nrow(subgroup_changes)]
  
  # School-level variance changes
  variance_change <- round((overall_changes$SCHOOL_RATE_SD - (overall_changes$SCHOOL_RATE_SD - overall_changes$SCHOOL_RATE_SD_CHANGE)) * 100, 2)
  q25_change <- round(overall_changes$SCHOOL_RATE_Q25_CHANGE * 100, 1)
  q75_change <- round(overall_changes$SCHOOL_RATE_Q75_CHANGE * 100, 1)
  
  cat("From", prev_year, "to", current_year_yoy, ", total district enrollment changed from", prev_enrollment, "to", current_enrollment, "students.")
  cat("\nThis represents an enrollment change of", enrollment_change, "students.")
  cat("\nThe number of chronically absent students changed from", prev_chronic_count, "in", prev_year, "to", current_chronic_count, "in", current_year_yoy, ".")
  cat("\nThis represents a change of", chronic_count_change, "chronically absent students.")
  
  cat("\n\nThe district chronic absence rate changed from", prev_rate, "% in", prev_year, "to", current_rate, "% in", current_year_yoy, ".")
  cat("\nThis represents a change of", rate_change, "percentage points.")
  
  if (rate_change > 0) {
    cat("\nThe district experienced an increase in chronic absence rate.")
  } else if (rate_change < 0) {
    cat("\nThe district experienced a decrease in chronic absence rate.")
  } else {
    cat("\nThe district chronic absence rate remained unchanged.")
  }
  
  cat("\n\nAmong demographic subgroups,", largest_increase_subgroup, "students had the largest rate increase at", largest_increase_change, "percentage points.")
  cat("\n", largest_decrease_subgroup, "students had the smallest rate change at", largest_decrease_change, "percentage points.")
  
  # Calculate range of subgroup changes
  subgroup_range <- max(subgroup_changes$rate_change_pp) - min(subgroup_changes$rate_change_pp)
  cat("\nSubgroup rate changes ranged from", min(subgroup_changes$rate_change_pp), "to", max(subgroup_changes$rate_change_pp), "percentage points.")
  cat("\nThis represents a", round(subgroup_range, 1), "percentage point spread in rate changes across demographic groups.")
  
  if (!is.na(variance_change)) {
    cat("\n\nThe standard deviation of school chronic absence rates changed by", abs(variance_change), "percentage points.")
    if (variance_change > 0) {
      cat("\nThis indicates increased variation among schools.")
    } else if (variance_change < 0) {
      cat("\nThis indicates decreased variation among schools.")
    }
  }
  
  if (!is.na(q25_change) && !is.na(q75_change)) {
    cat("\nThe first quartile of school rates changed by", q25_change, "percentage points.")
    cat("\nThe third quartile of school rates changed by", q75_change, "percentage points.")
    
    quartile_gap_change <- q75_change - q25_change
    if (quartile_gap_change > 0) {
      cat("\nThe gap between high and low performing schools increased by", round(quartile_gap_change, 1), "percentage points.")
    } else if (quartile_gap_change < 0) {
      cat("\nThe gap between high and low performing schools decreased by", round(abs(quartile_gap_change), 1), "percentage points.")
    }
  }
  
} else {
  cat("No year-over-year comparison data available for this district.")
  cat("\nThis indicates that the district has data for only one year in the dataset.")
}
```
