# School-Level Patterns

## Distribution of Chronic Absence Rates

```{r school-distribution-calcs, echo=FALSE}
# Get current year school data
school_current <- params$school_data %>% 
  filter(YEAR == current_year, SUBGROUP == "ALL") %>%
  arrange(CHRONIC_RATE)

n_schools_total <- nrow(school_current)
lowest_school_rate <- round(min(school_current$CHRONIC_RATE) * 100, 1)
highest_school_rate <- round(max(school_current$CHRONIC_RATE) * 100, 1)
rate_range <- highest_school_rate - lowest_school_rate
median_school_rate <- round(median(school_current$CHRONIC_RATE) * 100, 1)
mean_school_rate <- round(mean(school_current$CHRONIC_RATE) * 100, 1)

# Quartile calculations
q1_rate <- round(quantile(school_current$CHRONIC_RATE, 0.25) * 100, 1)
q3_rate <- round(quantile(school_current$CHRONIC_RATE, 0.75) * 100, 1)
iqr_range <- q3_rate - q1_rate

# Schools in different performance bands
schools_below_10 <- sum(school_current$CHRONIC_RATE < 0.10)
schools_10_to_20 <- sum(school_current$CHRONIC_RATE >= 0.10 & school_current$CHRONIC_RATE < 0.20)
schools_20_to_30 <- sum(school_current$CHRONIC_RATE >= 0.20 & school_current$CHRONIC_RATE < 0.30)
schools_above_30 <- sum(school_current$CHRONIC_RATE >= 0.30)

# Standard deviation calculation
school_rate_sd <- round(sd(school_current$CHRONIC_RATE) * 100, 2)

# Get school names for extremes
lowest_school_name <- school_current$SCHOOL_NAME[1]
highest_school_name <- school_current$SCHOOL_NAME[nrow(school_current)]

# Average enrollment per school
avg_enrollment_per_school <- round(mean(school_current$ENROLLMENT), 0)
total_district_enrollment <- sum(school_current$ENROLLMENT)
total_district_chronic <- sum(school_current$CHRONIC_COUNT)

# Schools above and below district average
district_avg_rate <- round(mean(params$district_data %>% filter(YEAR == current_year, SUBGROUP == "ALL") %>% pull(CHRONIC_RATE)) * 100, 1)
schools_above_district_avg <- sum(school_current$CHRONIC_RATE * 100 > district_avg_rate)
schools_below_district_avg <- sum(school_current$CHRONIC_RATE * 100 < district_avg_rate)
schools_at_district_avg <- sum(round(school_current$CHRONIC_RATE * 100, 1) == district_avg_rate)
```

Chronic absence rates in `r current_year` ranged from `r lowest_school_rate`% at `r lowest_school_name` to `r highest_school_rate`% at `r highest_school_name`.
This represents a range of `r rate_range` percentage points across the district's `r n_schools_total` schools.
The median school chronic absence rate was `r median_school_rate`% while the mean rate was `r mean_school_rate`%.
The standard deviation of school rates was `r school_rate_sd` percentage points.

The first quartile of schools had chronic absence rates at or below `r q1_rate`%.
The third quartile threshold was `r q3_rate`%.
The interquartile range spanned `r iqr_range` percentage points.
Half of all schools had chronic absence rates between these quartile boundaries.

Schools were distributed across performance levels as follows: `r schools_below_10` schools had rates below 10%, `r schools_10_to_20` schools had rates between 10% and 20%, `r schools_20_to_30` schools had rates between 20% and 30%, and `r schools_above_30` schools had rates above 30%.
The district average chronic absence rate was `r district_avg_rate`%.
`r schools_above_district_avg` schools exceeded this average while `r schools_below_district_avg` schools fell below the average.
`r if(schools_at_district_avg > 0) paste(schools_at_district_avg, "schools matched the district average exactly.") else "No schools matched the district average exactly."`

Schools enrolled an average of `r avg_enrollment_per_school` students each.
The total district enrollment of `r total_district_enrollment` students included `r total_district_chronic` chronically absent students.
Individual school enrollment ranged from `r min(school_current$ENROLLMENT)` to `r max(school_current$ENROLLMENT)` students.
The largest schools served `r round(max(school_current$ENROLLMENT) / min(school_current$ENROLLMENT), 1)` times as many students as the smallest schools.

### School Rate Standard Deviations by Subgroup (Percentage Points)

```{r school-variance-table, echo=FALSE}
# Create school variance table with years as columns - SD only
school_variance <- params$district_data %>%
  filter(!is.na(SCHOOL_RATE_SD)) %>%
  select(YEAR, SUBGROUP, SCHOOL_RATE_SD) %>%
  mutate(SD = round(SCHOOL_RATE_SD * 100, 2)) %>%
  select(YEAR, SUBGROUP, SD) %>%
  arrange(match(SUBGROUP, c("ALL", "MALE", "FEMALE", "HISPANIC", "WHITE", "ASIAN", "BLACK", "ELL", "DISADVANTAGE", "DISABILITY")), YEAR) %>%
  tidyr::pivot_wider(names_from = YEAR, values_from = SD)

kable(school_variance,
      caption = "Standard deviation of school chronic absence rates by subgroup across years.",
      align = c("l", rep("r", ncol(school_variance)-1)))
```

## School-by-School Breakdown

```{r school-breakdown-calcs, echo=FALSE}
# Get top 10 schools by chronic absence count in current year
top_chronic_schools <- params$school_data %>%
  filter(YEAR == current_year, SUBGROUP == "ALL") %>%
  arrange(desc(CHRONIC_COUNT)) %>%
  slice(1:10)

top_school_names <- paste(top_chronic_schools$SCHOOL_NAME, collapse = ", ")
top_school_chronic_counts <- top_chronic_schools$CHRONIC_COUNT
top_school_rates <- round(top_chronic_schools$CHRONIC_RATE * 100, 1)
top_school_enrollments <- top_chronic_schools$ENROLLMENT

# School with most chronic absence
highest_count_school <- top_chronic_schools$SCHOOL_NAME[1]
highest_count_value <- top_chronic_schools$CHRONIC_COUNT[1]
highest_count_rate <- round(top_chronic_schools$CHRONIC_RATE[1] * 100, 1)
highest_count_enrollment <- top_chronic_schools$ENROLLMENT[1]

# Calculate what proportion of district chronic absence comes from top schools
top_5_chronic_total <- sum(top_chronic_schools$CHRONIC_COUNT[1:5])
district_chronic_total <- sum(params$school_data %>% filter(YEAR == current_year, SUBGROUP == "ALL") %>% pull(CHRONIC_COUNT))
top_5_proportion <- round((top_5_chronic_total / district_chronic_total) * 100, 1)

top_10_chronic_total <- sum(top_chronic_schools$CHRONIC_COUNT)
top_10_proportion <- round((top_10_chronic_total / district_chronic_total) * 100, 1)

# Range of chronic counts in top 10
min_top_10_count <- min(top_chronic_schools$CHRONIC_COUNT)
max_top_10_count <- max(top_chronic_schools$CHRONIC_COUNT)
```

The school with the highest number of chronically absent students was `r highest_count_school` with `r highest_count_value` chronically absent students out of `r highest_count_enrollment` enrolled.
This represents a chronic absence rate of `r highest_count_rate`% at this school.
The top 10 schools by chronic absence count accounted for `r top_10_chronic_total` chronically absent students, representing `r top_10_proportion`% of all district chronic absence.
The top 5 schools alone accounted for `r top_5_chronic_total` chronically absent students, or `r top_5_proportion`% of the district total.

Among the top 10 schools by chronic absence count, the number of chronically absent students ranged from `r min_top_10_count` to `r max_top_10_count`.
These schools had enrollments ranging from `r min(top_chronic_schools$ENROLLMENT)` to `r max(top_chronic_schools$ENROLLMENT)` students.
Chronic absence rates at these schools ranged from `r min(top_school_rates)`% to `r max(top_school_rates)`%.
The average chronic absence rate among these 10 schools was `r round(mean(top_school_rates), 1)`%.

### Top 5 School Rankings Within District

```{r school-rankings-table, echo=FALSE}
# Create school rankings table - top 5 schools by chronic absence rate
school_rankings <- params$school_data %>%
  filter(YEAR == current_year, SUBGROUP == "ALL") %>%
  arrange(desc(CHRONIC_RATE)) %>%
  slice(1:5) %>%
  select(SCHOOL_NAME, ENROLLMENT, CHRONIC_COUNT, CHRONIC_RATE) %>%
  mutate(CHRONIC_RATE = round(CHRONIC_RATE * 100, 1))

kable(school_rankings,
      col.names = c("School Name", "Enrollment", "Chronic Count", "Rate (%)"),
      caption = "Top 5 schools by chronic absence rate within the district.",
      align = c("l", "r", "r", "r"))
```
