---
title: "Student Engagement Analysis: Assessment Participation and Chronic Absenteeism"
subtitle: "Multi-Year Trends for Data Quality and Accountability"
format: 
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
  docx:
    toc: false
params:
  data_file: "~/bueller_project/data_files/rds/merged_data.rds"
  calc_functions_path: "~/bueller_project/bueller/R/bueller_calc.R"
  output_year: 2023
  min_years: 3
  min_enrollment: 30
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)
library(ggplot2)

# Load and prepare data
data(df)

# Filter out County Offices of Education
county_mask <- !grepl("County Office of Education|County Department of Education", 
                     df$DISTRICT_NAME, ignore.case = TRUE)
df <- df[county_mask, ]

# Calculate district-level data using functions
district_data <- calc_district_metrics(df)

# Calculate school variation data using functions
school_variation <- calc_school_variation(df)

# Prepare district trends data using functions
district_trends <- prepare_district_trends(district_data, params$min_years)

# Filter by minimum enrollment
district_trends <- district_trends[district_trends$students_enrolled >= params$min_enrollment, ]

# Calculate key metrics for narrative
current_year <- max(district_trends$YEAR, na.rm = TRUE)
first_year <- min(district_trends$YEAR, na.rm = TRUE)

# Get current year data
current_data <- district_trends[
  district_trends$YEAR == current_year & 
  district_trends$DISAGGREGATED_GROUP == "All_Students", 
]

# Calculate summary statistics
ela_data <- current_data[current_data$SUBJECT == "ELA", ]
math_data <- current_data[current_data$SUBJECT == "MATH", ]

summary_stats <- rbind(
  data.frame(
    SUBJECT = "ELA",
    total_districts = nrow(ela_data),
    median_participation = round(median(ela_data$participation_rate, na.rm = TRUE), 2),
    median_absenteeism = round(median(ela_data$chronic_absent_rate, na.rm = TRUE), 2),
    pct_below_95 = round(mean(ela_data$participation_rate < 95, na.rm = TRUE) * 100, 2),
    pct_above_20 = round(mean(ela_data$chronic_absent_rate > 20, na.rm = TRUE) * 100, 2),
    total_enrollment = sum(ela_data$students_enrolled, na.rm = TRUE)
  ),
  data.frame(
    SUBJECT = "MATH",
    total_districts = nrow(math_data),
    median_participation = round(median(math_data$participation_rate, na.rm = TRUE), 2),
    median_absenteeism = round(median(math_data$chronic_absent_rate, na.rm = TRUE), 2),
    pct_below_95 = round(mean(math_data$participation_rate < 95, na.rm = TRUE) * 100, 2),
    pct_above_20 = round(mean(math_data$chronic_absent_rate > 20, na.rm = TRUE) * 100, 2),
    total_enrollment = sum(math_data$students_enrolled, na.rm = TRUE)
  )
)

# Extract ELA and MATH summaries
ela_summary <- summary_stats[summary_stats$SUBJECT == "ELA", ]
math_summary <- summary_stats[summary_stats$SUBJECT == "MATH", ]

# Year-over-year analysis
all_students_data <- district_trends[district_trends$DISAGGREGATED_GROUP == "All_Students", ]

yearly_trends <- aggregate(
  cbind(participation_rate, chronic_absent_rate) ~ YEAR + SUBJECT,
  data = all_students_data,
  FUN = function(x) {
    c(
      districts = length(x),
      median_participation = round(median(x, na.rm = TRUE), 2),
      median_absenteeism = round(median(x, na.rm = TRUE), 2),
      pct_below_95 = round(mean(x < 95, na.rm = TRUE) * 100, 2),
      pct_above_20 = round(mean(x > 20, na.rm = TRUE) * 100, 2)
    )
  }
)

# Flatten the aggregate results
yearly_trends <- data.frame(
  yearly_trends[1:2],
  yearly_trends$participation_rate[, c(1:2, 4)],  # districts, median, pct_below_95
  yearly_trends$chronic_absent_rate[, c(3, 5)]    # median_absenteeism, pct_above_20
)
names(yearly_trends) <- c("YEAR", "SUBJECT", "districts", "median_participation", 
                         "pct_below_95", "median_absenteeism", "pct_above_20")

# School-level driver analysis - PARTICIPATION by subject
current_year_data <- df[
  df$YEAR == current_year & 
  df$DISAGGREGATED_GROUP == "All_Students", 
]

participation_drivers <- aggregate(
  PARTICIPATION_RATE ~ COUNTY_ID + DISTRICT_ID + COUNTY_NAME + DISTRICT_NAME + SUBJECT,
  data = current_year_data,
  FUN = function(x) {
    if (length(x) >= 3) {
      c(
        total_schools = length(x),
        schools_below_90_participation = sum(x < 90, na.rm = TRUE),
        min_participation = min(x, na.rm = TRUE),
        max_participation = max(x, na.rm = TRUE),
        participation_range = max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
      )
    } else {
      c(0, 0, 0, 0, 0)
    }
  }
)

# Flatten participation results
if (nrow(participation_drivers) > 0) {
  part_matrix <- participation_drivers$PARTICIPATION_RATE
  participation_drivers <- data.frame(
    participation_drivers[1:5],
    total_schools = part_matrix[, 1],
    schools_below_90_participation = part_matrix[, 2],
    min_participation = part_matrix[, 3],
    max_participation = part_matrix[, 4],
    participation_range = part_matrix[, 5]
  )
  
  # Filter and calculate percentages
  participation_drivers <- participation_drivers[participation_drivers$total_schools >= 3, ]
  participation_drivers$pct_schools_low_participation <- round(
    participation_drivers$schools_below_90_participation / participation_drivers$total_schools * 100, 2
  )
  participation_drivers$driven_by_few_schools_participation <- 
    participation_drivers$schools_below_90_participation > 0 & 
    participation_drivers$pct_schools_low_participation <= 33
}

# School-level driver analysis - ABSENTEEISM (not by subject)
# First get unique schools per district
school_absent_data <- aggregate(
  CHRONIC_ABSENT_RATE ~ COUNTY_ID + DISTRICT_ID + SCHOOL_ID,
  data = current_year_data,
  FUN = function(x) x[1]
)

# Add district names
district_info <- unique(current_year_data[c("COUNTY_ID", "DISTRICT_ID", "COUNTY_NAME", "DISTRICT_NAME")])
school_key <- paste(school_absent_data$COUNTY_ID, school_absent_data$DISTRICT_ID)
district_key <- paste(district_info$COUNTY_ID, district_info$DISTRICT_ID)
match_idx <- match(school_key, district_key)
school_absent_data$COUNTY_NAME <- district_info$COUNTY_NAME[match_idx]
school_absent_data$DISTRICT_NAME <- district_info$DISTRICT_NAME[match_idx]

# Aggregate by district
absenteeism_drivers <- aggregate(
  CHRONIC_ABSENT_RATE ~ COUNTY_ID + DISTRICT_ID + COUNTY_NAME + DISTRICT_NAME,
  data = school_absent_data,
  FUN = function(x) {
    if (length(x) >= 3) {
      c(
        total_schools = length(x),
        schools_above_30_absenteeism = sum(x > 30, na.rm = TRUE),
        min_absenteeism = min(x, na.rm = TRUE),
        max_absenteeism = max(x, na.rm = TRUE),
        absenteeism_range = max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
      )
    } else {
      c(0, 0, 0, 0, 0)
    }
  }
)

# Flatten absenteeism results
if (nrow(absenteeism_drivers) > 0) {
  absent_matrix <- absenteeism_drivers$CHRONIC_ABSENT_RATE
  absenteeism_drivers <- data.frame(
    absenteeism_drivers[1:4],
    total_schools = absent_matrix[, 1],
    schools_above_30_absenteeism = absent_matrix[, 2],
    min_absenteeism = absent_matrix[, 3],
    max_absenteeism = absent_matrix[, 4],
    absenteeism_range = absent_matrix[, 5]
  )
  
  # Filter and calculate percentages
  absenteeism_drivers <- absenteeism_drivers[absenteeism_drivers$total_schools >= 3, ]
  absenteeism_drivers$pct_schools_high_absenteeism <- round(
    absenteeism_drivers$schools_above_30_absenteeism / absenteeism_drivers$total_schools * 100, 2
  )
  absenteeism_drivers$driven_by_few_schools_absenteeism <- 
    absenteeism_drivers$schools_above_30_absenteeism > 0 & 
    absenteeism_drivers$pct_schools_high_absenteeism <= 33
}

# Count districts driven by few schools
ela_driven_participation <- if (exists("participation_drivers") && nrow(participation_drivers) > 0) {
  sum(participation_drivers$SUBJECT == "ELA" & participation_drivers$driven_by_few_schools_participation, na.rm = TRUE)
} else { 0 }

math_driven_participation <- if (exists("participation_drivers") && nrow(participation_drivers) > 0) {
  sum(participation_drivers$SUBJECT == "MATH" & participation_drivers$driven_by_few_schools_participation, na.rm = TRUE)
} else { 0 }

districts_driven_absenteeism <- if (exists("absenteeism_drivers") && nrow(absenteeism_drivers) > 0) {
  sum(absenteeism_drivers$driven_by_few_schools_absenteeism, na.rm = TRUE)
} else { 0 }

# Student group analysis
group_data <- district_trends[
  district_trends$YEAR == current_year & 
  district_trends$DISAGGREGATED_GROUP != "All_Students", 
]

all_students_current <- district_trends[
  district_trends$YEAR == current_year & 
  district_trends$DISAGGREGATED_GROUP == "All_Students", 
  c("COUNTY_ID", "DISTRICT_ID", "SUBJECT", "participation_rate", "chronic_absent_rate")
]

# Merge district averages
group_key <- paste(group_data$COUNTY_ID, group_data$DISTRICT_ID, group_data$SUBJECT)
all_key <- paste(all_students_current$COUNTY_ID, all_students_current$DISTRICT_ID, all_students_current$SUBJECT)
match_indices <- match(group_key, all_key)

group_data$district_participation <- all_students_current$participation_rate[match_indices]
group_data$district_absenteeism <- all_students_current$chronic_absent_rate[match_indices]
group_data$participation_gap <- group_data$district_participation - group_data$participation_rate
group_data$absenteeism_gap <- group_data$chronic_absent_rate - group_data$district_absenteeism

# Filter by enrollment
group_analysis <- group_data[group_data$students_enrolled >= 10, ]

# Find largest gaps
if (nrow(group_analysis) > 0) {
  largest_participation_gap <- group_analysis[order(-group_analysis$participation_gap, na.last = TRUE), ][1, ]
  largest_absenteeism_gap <- group_analysis[order(-group_analysis$absenteeism_gap, na.last = TRUE), ][1, ]
  significant_participation_gaps <- sum(group_analysis$participation_gap >= 15, na.rm = TRUE)
  significant_absenteeism_gaps <- sum(group_analysis$absenteeism_gap >= 15, na.rm = TRUE)
} else {
  largest_participation_gap <- data.frame(
    DISAGGREGATED_GROUP = "None", DISTRICT_NAME = "None",
    participation_rate = 0, district_participation = 0, participation_gap = 0
  )
  largest_absenteeism_gap <- data.frame(
    DISAGGREGATED_GROUP = "None", DISTRICT_NAME = "None",
    chronic_absent_rate = 0, district_absenteeism = 0, absenteeism_gap = 0
  )
  significant_participation_gaps <- 0
  significant_absenteeism_gaps <- 0
}

# Most extreme districts
ela_current <- current_data[current_data$SUBJECT == "ELA", ]
if (nrow(ela_current) > 0) {
  worst_participation <- ela_current[order(ela_current$participation_rate), ][1, ]
  best_participation <- ela_current[order(-ela_current$participation_rate), ][1, ]
  worst_absenteeism <- ela_current[order(-ela_current$chronic_absent_rate), ][1, ]
  best_absenteeism <- ela_current[order(ela_current$chronic_absent_rate), ][1, ]
} else {
  default_row <- data.frame(participation_rate = 0, chronic_absent_rate = 0, DISTRICT_NAME = "None")
  worst_participation <- best_participation <- worst_absenteeism <- best_absenteeism <- default_row
}

# Validate key variables exist
stopifnot(
  exists("ela_driven_participation"), is.numeric(ela_driven_participation),
  exists("math_driven_participation"), is.numeric(math_driven_participation),
  exists("districts_driven_absenteeism"), is.numeric(districts_driven_absenteeism),
  nrow(largest_participation_gap) == 1,
  nrow(largest_absenteeism_gap) == 1,
  nrow(worst_participation) == 1,
  nrow(worst_absenteeism) == 1,
  nrow(best_participation) == 1,
  nrow(best_absenteeism) == 1
)
```

---

## What is the current state of assessment participation and chronic absenteeism across California districts?

**This analysis examines student engagement patterns across `r ela_summary$total_districts` California districts serving `r format(ela_summary$total_enrollment, big.mark = ",")` students during the `r first_year`-`r current_year` period.** **Current statewide performance shows median district participation rates of `r ela_summary$median_participation`% (ELA) and `r math_summary$median_participation`% (Math), while chronic absenteeism has reached median levels of `r ela_summary$median_absenteeism`%.**

**Assessment participation directly impacts the validity of accountability determinations and federal compliance status.** **The statewide landscape shows considerable variation, with district participation rates ranging from `r worst_participation$participation_rate`% (`r worst_participation$DISTRICT_NAME`) to `r best_participation$participation_rate`% (`r best_participation$DISTRICT_NAME`), demonstrating a `r round(best_participation$participation_rate - worst_participation$participation_rate, 2)` percentage point range across districts.**

**Notably, `r ela_summary$pct_below_95`% of districts fall below the federal 95% participation threshold, and `r ela_summary$pct_above_20`% exceed the 20% chronic absenteeism benchmark.** **Current data shows district absenteeism rates ranging from `r best_absenteeism$chronic_absent_rate`% (`r best_absenteeism$DISTRICT_NAME`) to `r worst_absenteeism$chronic_absent_rate`% (`r worst_absenteeism$DISTRICT_NAME`), representing a `r round(worst_absenteeism$chronic_absent_rate - best_absenteeism$chronic_absent_rate, 2)` percentage point range across the state.**

---

## How have participation and absenteeism patterns changed over time?

### Multi-Year Participation Trends

```{r participation-trends-table}
# Create pivot table manually
years <- sort(unique(yearly_trends$YEAR))
ela_trends <- yearly_trends[yearly_trends$SUBJECT == "ELA", ]
math_trends <- yearly_trends[yearly_trends$SUBJECT == "MATH", ]

participation_trends_table <- data.frame(
  YEAR = years,
  median_participation_ELA = ela_trends$median_participation[match(years, ela_trends$YEAR)],
  median_participation_MATH = math_trends$median_participation[match(years, math_trends$YEAR)],
  pct_below_95_ELA = ela_trends$pct_below_95[match(years, ela_trends$YEAR)],
  pct_below_95_MATH = math_trends$pct_below_95[match(years, math_trends$YEAR)]
)

kable(participation_trends_table,
      col.names = c("Year", "ELA Median %", "Math Median %", "ELA % Below 95%", "Math % Below 95%"),
      caption = "District Assessment Participation Trends by Subject and Year")
```

**The trend data reveals that participation rates experienced their most significant disruption in 2020, with median participation dropping to unprecedented levels before beginning recovery in subsequent years.** **The federal compliance picture shows `r ela_trends$pct_below_95[ela_trends$YEAR == current_year]`% of districts currently failing to meet the 95% threshold in ELA, compared to `r ela_trends$pct_below_95[ela_trends$YEAR == first_year]`% in `r first_year`.**

### Multi-Year Absenteeism Trends

```{r absenteeism-trends-table}
absenteeism_trends_table <- data.frame(
  YEAR = years,
  median_absenteeism_ELA = ela_trends$median_absenteeism[match(years, ela_trends$YEAR)],
  median_absenteeism_MATH = math_trends$median_absenteeism[match(years, math_trends$YEAR)],
  pct_above_20_ELA = ela_trends$pct_above_20[match(years, ela_trends$YEAR)],
  pct_above_20_MATH = math_trends$pct_above_20[match(years, math_trends$YEAR)]
)

kable(absenteeism_trends_table,
      col.names = c("Year", "ELA Median %", "Math Median %", "ELA % Above 20%", "Math % Above 20%"),
      caption = "District Chronic Absenteeism Trends by Subject and Year")
```

**Chronic absenteeism patterns show dramatic increases during the 2020-2021 period, with median district rates reaching peak levels before beginning to moderate in recent years.** **The proportion of districts exceeding the 20% threshold increased substantially during this period, with `r ela_trends$pct_above_20[ela_trends$YEAR == current_year]`% currently above this benchmark compared to `r ela_trends$pct_above_20[ela_trends$YEAR == first_year]`% in `r first_year`.**

---

## Are engagement challenges concentrated in specific schools within districts or distributed system-wide?

### School-Level Participation Drivers

```{r participation-drivers-plot, fig.height=6, fig.width=12}
if (exists("participation_drivers") && nrow(participation_drivers) > 0) {
  # Filter to districts with at least 5 schools and get top 10 per subject
  ela_viz <- participation_drivers[
    participation_drivers$total_schools >= 5 & 
    participation_drivers$SUBJECT == "ELA", 
  ]
  math_viz <- participation_drivers[
    participation_drivers$total_schools >= 5 & 
    participation_drivers$SUBJECT == "MATH", 
  ]
  
  ela_top <- ela_viz[order(-ela_viz$participation_range), ][1:min(10, nrow(ela_viz)), ]
  math_top <- math_viz[order(-math_viz$participation_range), ][1:min(10, nrow(math_viz)), ]
  
  participation_viz_data <- rbind(ela_top, math_top)
  participation_viz_data$district_label <- paste(
    participation_viz_data$DISTRICT_NAME, 
    participation_viz_data$COUNTY_NAME, 
    sep = "\n"
  )
  
  ggplot(participation_viz_data, aes(x = reorder(district_label, participation_range))) +
    geom_col(aes(y = participation_range, fill = SUBJECT), alpha = 0.7) +
    geom_text(aes(y = participation_range / 2, 
                  label = paste0(schools_below_90_participation, "/", total_schools)), 
              size = 2.5, hjust = 0.5, vjust = 0.5, color = "white", fontface = "bold") +
    scale_fill_manual(values = c("ELA" = "steelblue", "MATH" = "darkgreen")) +
    coord_flip() +
    facet_wrap(~ SUBJECT, scales = "free", ncol = 2) +
    labs(title = "Districts with Largest School-to-School Participation Variation by Subject",
         subtitle = "Top 10 districts by participation range within each subject",
         x = "District", 
         y = "Participation Rate Range (percentage points)",
         caption = "Labels show number of schools below 90% participation") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 7),
          legend.position = "none")
}
```

**This analysis identifies districts where participation challenges are concentrated in specific schools rather than distributed system-wide.** **Among districts with at least 5 schools, `r ela_driven_participation` show ELA participation problems driven by one-third or fewer of their schools, while `r math_driven_participation` show similar patterns in MATH, suggesting that targeted interventions at the school level could significantly improve district-wide compliance rates.**

### School-Level Absenteeism Drivers

```{r absenteeism-drivers-plot, fig.height=6, fig.width=10}
if (exists("absenteeism_drivers") && nrow(absenteeism_drivers) > 0) {
  absenteeism_viz_data <- absenteeism_drivers[absenteeism_drivers$total_schools >= 5, ]
  absenteeism_viz_data <- absenteeism_viz_data[
    order(-absenteeism_viz_data$absenteeism_range), 
  ][1:min(15, nrow(absenteeism_viz_data)), ]
  
  ggplot(absenteeism_viz_data, aes(x = reorder(paste(DISTRICT_NAME, COUNTY_NAME, sep = "\n"), absenteeism_range))) +
    geom_col(aes(y = absenteeism_range), fill = "coral", alpha = 0.7) +
    geom_text(aes(y = absenteeism_range / 2, 
                  label = paste0(schools_above_30_absenteeism, "/", total_schools)), 
              size = 3, hjust = 0.5, vjust = 0.5, color = "white", fontface = "bold") +
    coord_flip() +
    labs(title = "Districts with Largest School-to-School Absenteeism Variation",
         subtitle = "Range between highest and lowest performing schools within district",
         x = "District", 
         y = "Chronic Absenteeism Range (percentage points)",
         caption = "Labels show number of schools above 30% chronic absenteeism") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 8))
}
```

**Similar to participation patterns, chronic absenteeism challenges are often concentrated within specific schools rather than being district-wide phenomena.** **Analysis reveals that `r districts_driven_absenteeism` districts have absenteeism concerns driven by one-third or fewer of their schools, indicating that school-specific factors—such as leadership, climate, or targeted populations—may be primary drivers of attendance challenges.**

**A critical finding is that participation and absenteeism challenges are often concentrated in a small number of schools within districts—`r ela_driven_participation` districts have ELA participation issues driven by one-third or fewer of their schools, `r math_driven_participation` districts show similar patterns for MATH participation, while `r districts_driven_absenteeism` districts show concentration patterns for chronic absenteeism.** This concentration suggests that targeted school-level interventions may be more effective than district-wide approaches in many cases.

---

## Which student groups show the largest engagement gaps compared to their district averages?

**Disaggregated data analysis reveals substantial variation in engagement patterns across student groups, with some populations experiencing significantly lower participation rates and higher absenteeism than their district averages.** **The largest recorded participation gap shows `r largest_participation_gap$DISAGGREGATED_GROUP` students in `r largest_participation_gap$DISTRICT_NAME` participating at `r round(largest_participation_gap$participation_rate, 2)`%, compared to a district average of `r round(largest_participation_gap$district_participation, 2)`%—a difference of `r round(largest_participation_gap$participation_gap, 2)` percentage points.**

**For chronic absenteeism, the most substantial gap appears among `r largest_absenteeism_gap$DISAGGREGATED_GROUP` students in `r largest_absenteeism_gap$DISTRICT_NAME`, with an absenteeism rate of `r round(largest_absenteeism_gap$chronic_absent_rate, 2)`% compared to the district average of `r round(largest_absenteeism_gap$district_absenteeism, 2)`%—representing a `r round(largest_absenteeism_gap$absenteeism_gap, 2)` percentage point difference.**

**Across all districts and student groups, `r significant_participation_gaps` instances show participation gaps of 15 percentage points or more, while `r significant_absenteeism_gaps` instances demonstrate similar magnitude gaps in chronic absenteeism.** These patterns suggest that engagement challenges disproportionately affect specific student populations, with implications for both educational equity and data quality in accountability systems.

---

## What are the implications for data quality and intervention strategies?

**The concentration of engagement challenges within specific schools and student groups has important implications for data interpretation and intervention planning.** **When participation or absenteeism problems are driven by a small number of schools within a district, district-wide averages may mask the severity of issues at particular sites while simultaneously understating the performance of schools that are functioning effectively.**

**The school-level concentration patterns identified in this analysis suggest that `r ela_driven_participation + math_driven_participation + districts_driven_absenteeism` district-metric combinations could benefit from targeted, school-specific interventions rather than broad district-wide approaches.** This finding is particularly relevant for states developing technical assistance and support strategies, as resources may be more effectively deployed when focused on the specific schools driving district-level challenges.

**Student group disparities further compound these data quality concerns, as substantial gaps in participation and attendance may result in certain populations being under-represented in accountability measures.** **The `r significant_participation_gaps + significant_absenteeism_gaps` instances of large engagement gaps identified in this analysis represent student groups whose academic performance may not be accurately reflected in state accountability systems due to low assessment participation or chronic absence patterns.**

*This analysis encompasses data from `r first_year` through `r current_year` across `r ela_summary$total_districts` districts meeting minimum enrollment criteria. District-level calculations exclude County Offices of Education and require minimum enrollment of `r params$min_enrollment` students for statistical reliability.*